---
title: "Frederick County School Board Race 2024"
output: html_document
date: "2026-06-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##LOAD LIBRARIES
```{r}
library(tidyverse)
library(janitor)
library(dplyr)
library(sf)
library(tidyr)
```

#importing, combining, cleaning data column names
```{r}
general1 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/GENERAL_11PG24_Prov_CVR_SBE_1.csv") |> 
  clean_names()
general2 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/GENERAL_11PG24_Prov_CVR_SBE.csv") |> 
  clean_names()

general_2024 <- rbind(general1, general2)

write_csv(general_2024, "general_2024.csv")



general_cleaned <- general |>
  select(cast_vote_record, precinct, seat_one = board_of_education_500001671, seat_two = x13, seat_three = x14)

seat_one <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_one)
seat_two <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_two)
seat_three <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_three)

combined <- union(seat_one, seat_two,)
combined <-union(combined, seat_three)


```
#focus on total of for each candidate by precinct, 

```{r}
 general_cleaned <- general |>
  select(cast_vote_record, precinct, ballot_style, seat_one = board_of_education_500001671, seat_two = x13, seat_three = x14)

seat_one <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_one)
seat_two <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_two)
seat_three <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_three)

combined <- union(seat_one, seat_two,)
combined <-union(combined, seat_three)



```

##QUESTIONS why are some undervotes and some are NA? should undervote be excluded or included? #getting this weird message, but the code still works
```{r}
precinct_vote_counts <- combined |>
   group_by(precinct, candidate)|>
   arrange(candidate)|>
  summarize(votes=n()) |>
  mutate(precinct= substring(precinct, 2))|>
  filter(
    !is.na(candidate),
    !candidate %in% c("undervote", "overvote", "or write-in:") #questionable
  ) |>
  group_by(precinct) |>
  arrange(desc(votes), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  ungroup()


precinct_vote_wide <- precinct_vote_counts |>
  pivot_wider(
    names_from = rank,
    values_from = c(candidate, votes),
  names_glue = "{.value}_{rank}")|>
 mutate(
    across(
      starts_with("candidate"),
      ~ trimws(gsub("\\s*\\([^)]*\\)", "",.x))
    ))

  write.csv(precinct_vote_wide, "precinct_vote_wide.csv", row.names = FALSE)


  #create a map to represent the percent of voters per precinct who voted for rep-backed candidates versus dem-backed candidates— easier to compare with kamala v trump


vote_percent_board <- precinct_vote_counts |>
  group_by(precinct)|>
  mutate(
    across(
            starts_with("candidate"),
               ~ trimws(gsub("\\s*\\([^)]*\\)", "",.x))),
    party =
          case_when(
              candidate == "Jaime Kiersten Brennan" ~ "*",
              candidate == "Colt Morningstar Black" ~ "*",
              candidate == "Janie Monier" ~ "^",
              candidate == "Josh Bokee" ~ "^",
              candidate == "Chad King Wilson, Sr." ~ "^",
              candidate == "Veronica D. Lowe"~ "^"))|>
        pivot_wider(
            names_from = rank,
            values_from = c(candidate, votes, party),
            names_glue = "{.value}_{rank}")|>
      mutate(
           total_votes = votes_1 + votes_2 + votes_3 + votes_4 + votes_5 + votes_6,
           percent_votes_1 = round((votes_1/total_votes)*100),
           percent_votes_2 = round((votes_2/total_votes)*100),
           percent_votes_3 = round((votes_3/total_votes)*100),
           percent_votes_4 = round((votes_4/total_votes)*100),
           percent_votes_5 = round((votes_5/total_votes)*100),
           percent_votes_6 = round((votes_6/total_votes)*100)
     
     )

print(vote_percent_board)

  write.csv(vote_percent_board, "vote_percent.csv", row.names = FALSE)


```



```{r}

```



##display by top candidates
top candidates:
winner:
Jaime Brennan and Colt Morningstar Black were two of the top three vote-getters in Frederick County. Brennan stepped down as chair of the county chapter of Moms for Liberty to run for school board seat;

Jaime Brennan             16.8% 	55,444 (REP/ Moms for liberty former chair)
Janie Monier              16.7% 	54,976  (DEM)
Colt Morningstar Black    16.2% 	53,403  (REP)
Josh Bokee	              16.1% 	53,190. (DEM)
Chad King Wilson Sr.	    14.8% 	48,860 (DEM)
Veronica Lowe             12.5% 	41,200 (DEM)


NOW PRECINCT MAP BY PRESIDENTIAL RESULTS
```{r}
general_president_clean <- general|>
  select(precinct, president=president_and_vice_president_of_the_united_states_500001630)|>
  group_by(precinct, president)|>
     arrange(president)|>
  summarize(votes=n()) |>
  mutate(precinct= substring(precinct, 2))|>
  filter(
    !is.na(president),
    !president %in% c("undervote", "overvote", "or write-in:") #questionable
  ) |>
  group_by(precinct) |>
  arrange(desc(votes), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  ungroup()|>
  mutate(
    across(
            starts_with("president"),
               ~ trimws(gsub("\\s*\\([^)]*\\)", "",.x))))|>
  pivot_wider(
    names_from = rank,
    values_from = c(president, votes),
  names_glue = "{.value}_{rank}")|>
      mutate(
           total_votes = votes_1 + votes_2 + votes_3 + votes_4 + votes_5,
           percent_votes_1 = round((votes_1/total_votes)*100),
           percent_votes_2 = round((votes_2/total_votes)*100),
           percent_votes_3 = round((votes_3/total_votes)*100),
           percent_votes_4 = round((votes_4/total_votes)*100),
           percent_votes_5 = round((votes_5/total_votes)*100)

     )
  

write.csv(general_president_clean, "president_vote.csv", row.names = FALSE)


harris_win <- general_president_clean |>
  group_by(precinct)|>
  sum(
     president_1 =="Harris-Walz")
  ) 
    



harris_win_versus_mol <- general_president_clean |>
  left_join( vote_percent_board, by = "precinct")|>
  select(precinct, president_1, candidate_1, candidate_2, candidate_3)|>
  filter(president_1=="Harris-Walz")|>
    filter(
    candidate_1 %in% c("Jaime Kiersten Brennan", "Colt Morningstar Black") | candidate_2 %in% c("Jaime Kiersten Brennan", "Colt Morningstar Black") | candidate_3 %in% c("Jaime Kiersten Brennan", "Colt Morningstar Black")
  )
          


```








  