---
title: "Draft story memo"
author: "Katelynn Winebrenner"
date: "June-1-2026"
format:
  html:
    df-print: kable
    theme: cosmo
---

katelynnwinebrenner\@gmail.com 240-362-4882 June 1, 2026

# 2024 Frederick County School Board Election

In the 2024 Frederick County School Board election, the Frederick County Democratic Party endorsed four candidates for three open at-large seats. This decision caused splintering within the party that led to conservative candidates -- Jamie Kiersten Brennan and Colt Morningstar Black -- to take two of the seats.

Brennan stepped down as chair of the county chapter of Moms for Liberty to run for school board seat. She won 16.8% of the total votes-- the highest in the county.

Janie Monier was the only candidate with support from the Democratic party to win, racking up 16.7% -- the second highest percentage of votes.

Black ran with conservative support. He won 16.2% of the votes -- the third top candidate.

Josh Bokee lost the race with 16.1% of the total votes. Chad King Wilson Sr. also lost with 14.8%. Veronica Lowe won the lowest 12.5%.

In rural precincts that voted overwhelmingly for President Donald Trump's second term, Colt and Brennan performed well.

In precincts within the City of Frederick, where former Vice President Kamala Harris won an overwhelming majority, candidate Josh Bokee and Veronica Lowe performed well. However, neither candidate earned enough votes to secure standing in the top three.

In precincts where Harris narrowly outperformed Trump, the vote for the Democratic-backed candidates split, allowing Brennan and Black to place in the top three.

Monier was the top Democrat-supported candidate in these precincts, earning her a seat on the board.

<iframe title="Democrat-backed school board candidates splintered precinct votes in 2024" aria-label="Choropleth map" id="datawrapper-chart-S5pgh" src="https://datawrapper.dwcdn.net/S5pgh/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height data-external="1">

</iframe>

```{=html}
<script type="text/javascript">(function(){function e(){window.addEventListener(`message`,function(e){if(e.data[`datawrapper-height`]!==void 0){var t=document.querySelectorAll(`iframe`);for(var n in e.data[`datawrapper-height`])for(var r=0,i;i=t[r];r++)if(i.contentWindow===e.source){var a=e.data[`datawrapper-height`][n]+`px`;i.style.height=a}}})}e()})();</script>
```

<iframe title="Harris took Frederick County, but Trump wasn&#39;t far behind" aria-label="Choropleth map" id="datawrapper-chart-bakfo" src="https://datawrapper.dwcdn.net/bakfo/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height data-external="1">

</iframe>

## Tables (Heads of my precinct data analysis)

```{r}
#| include: false
knitr::opts_chunk$set(echo = TRUE)


library(tidyverse)
library(janitor)
library(dplyr)
library(sf)
library(tidyr)
library(gt)


general1 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/GENERAL_11PG24_Prov_CVR_SBE_1.csv") |> 
  clean_names()

general2 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/GENERAL_11PG24_Prov_CVR_SBE.csv") |> 
  clean_names()

#the two general files from 2024 should be combined. they are the same sheet, one was just limited to 99,999 
general <- rbind(general1, general2)


general_cleaned <- general |>
  select(cast_vote_record, precinct, ballot_style, seat_one = board_of_education_500001671, seat_two = x13, seat_three = x14)

seat_one <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_one)
seat_two <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_two)
seat_three <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_three)

combined <- union(seat_one, seat_two,)
combined <-union(combined, seat_three)


precinct_vote_counts <- combined |>
  group_by(precinct, candidate)|>
  arrange(candidate)|>
  summarize(votes=n()) |>
  mutate(precinct= substring(precinct, 2))|>
  filter(
    !is.na(candidate),
    !candidate %in% c("undervote", "overvote", "or write-in:") #questionable
  ) |>
  group_by(precinct) |>
  arrange(desc(votes), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  ungroup()


precinct_vote_wide <- precinct_vote_counts |>
  pivot_wider(
    names_from = rank,
    values_from = c(candidate, votes),
    names_glue = "{.value}_{rank}")|>
  mutate(
    across(
      starts_with("candidate"),
      ~ trimws(gsub("\\s*\\([^)]*\\)", "",.x))
    ))

write.csv(precinct_vote_wide, "precinct_vote_wide.csv", row.names = FALSE)

#create a map to represent the percent of voters per precinct who voted for rep-backed candidates versus dem-backed candidates— easier to compare with kamala v trump

vote_percent_board <- precinct_vote_counts |>
  group_by(precinct)|>
  mutate(
    across(
      starts_with("candidate"),
      ~ trimws(gsub("\\s*\\([^)]*\\)", "",.x))),
    party =
      case_when(
        candidate == "Jaime Kiersten Brennan" ~ "*",
        candidate == "Colt Morningstar Black" ~ "*",
        candidate == "Janie Monier" ~ "^",
        candidate == "Josh Bokee" ~ "^",
        candidate == "Chad King Wilson, Sr." ~ "^",
        candidate == "Veronica D. Lowe"~ "^"))|>
  pivot_wider(
    names_from = rank,
    values_from = c(candidate, votes, party),
    names_glue = "{.value}_{rank}")|>
  mutate(
    total_votes = votes_1 + votes_2 + votes_3 + votes_4 + votes_5 + votes_6,
    percent_votes_1 = round((votes_1/total_votes)*100),
    percent_votes_2 = round((votes_2/total_votes)*100),
    percent_votes_3 = round((votes_3/total_votes)*100),
    percent_votes_4 = round((votes_4/total_votes)*100),
    percent_votes_5 = round((votes_5/total_votes)*100),
    percent_votes_6 = round((votes_6/total_votes)*100)
    
  )

write.csv(vote_percent_board, "vote_percent.csv", row.names = FALSE)




general_president_clean <- general|>
  select(precinct, president=president_and_vice_president_of_the_united_states_500001630)|>
  group_by(precinct, president)|>
  arrange(president)|>
  summarize(votes=n()) |>
  mutate(precinct= substring(precinct, 2))|>
  filter(
    !is.na(president),
    !president %in% c("undervote", "overvote", "or write-in:") #questionable
  ) |>
  group_by(precinct) |>
  arrange(desc(votes), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  ungroup()|>
  mutate(
    across(
      starts_with("president"),
      ~ trimws(gsub("\\s*\\([^)]*\\)", "",.x))))|>
  pivot_wider(
    names_from = rank,
    values_from = c(president, votes),
    names_glue = "{.value}_{rank}")|>
  mutate(
    total_votes = votes_1 + votes_2 + votes_3 + votes_4 + votes_5,
    percent_votes_1 = round((votes_1/total_votes)*100),
    percent_votes_2 = round((votes_2/total_votes)*100),
    percent_votes_3 = round((votes_3/total_votes)*100),
    percent_votes_4 = round((votes_4/total_votes)*100),
    percent_votes_5 = round((votes_5/total_votes)*100)
    
  )


```

```{r}
#| echo: false
vote_percent_board |>
  as_tibble() |>
  head()|>
  gt()

```

```{r}
#| echo: false
general_president_clean|>
  as_tibble() |>
  head()|>
  gt()
```

## Relevance

This story is relevant now because no one has mapped out these votes yet to identify where the splintering happened, and there is another upcoming school board election for the other four seats.

## Audience

My target audience is Frederick county parents and candidates.

## Sources

Some people I could interview for this story would be: Frederick Democrats Chair \| Mari Lee \| (301) 898-2513 \| info\@frederickdemocrats.org Moms for Liberty Chapter Chair \| Tiffany Noble \| Tiffany.noble0820\@gmail.com Frederick County PTA President \| Najlaa Richard \| president\@frederickpta.org School board candidate Chad Wilson, Sr. (lost in 2024, running again this season)\
It would be very interesting to go to these purple precincts to find someone who voted for Harris in 2024 but also voted for a conservative school board member to interview and to hear their decision making process.

## Potential impact

The potential impact of this story would be to reveal key areas where splintering occurred to inform candidates and voters where the discrepancies were to increase voter participation and civic engagement in these areas.

## Interest

This story was particularly surprising to me because it showcases the impact of partisanship within the county school board, which is technically considered nonpartisan. The extreme ideological differences between the candidates is also intriguing to me, particularly because the former Frederick County Moms of Liberty stepped down to run for a seat and won. I’m also interested in this story because it shows that, despite Frederick’s growth in recent years that have skewed its politics to the left, the conservative values in the county still remain strong.

## Previous coverage

From what I’ve seen, there has not been much coverage on this race in particular, other than generalized statewide coverage about the Democrats’ failures in school boards throughout the state from Maryland Matters.

Half of ‘extreme’ school board candidates targeted by Md. Democratic Party win their races - Maryland Matters

Md. Democratic Party targets 'extreme candidates ' in local school board races - Maryland Matters

## Visuals

I've already created a choropleth map of precincts that display where the splintering occurred. Comparing and contrasting that map to a map of 2024 election results would help contextualize these areas as leaning left or right as well. I'd like to include an extra html button in those maps to click back from one to the other as Adam Marton demonstrated for the final version or find a way to embed the maps side by side, so viewers can more easily compare presidential results to school board results.

I'd still like to create a bar graph of sorts to visualize the total votes each candidate received county wide. I will do the same with the the Harris v. Trump race.

## To learn

As for other information still to gather, I need to do more research on the previous election members, why those seats were open and what the 2024 campaign season was like. I’d also like to research what kind of decisions or changes the new conservative school board members have made since being elected to office and what new candidates are saying about them.

## To do and expected completion

I would like to have this done around the time of the primary on June 23. Interviews shouldn’t take very long to set up. I would also like to try to talk to some regular Frederick County parents, which could be a little more difficult as a newcomer. I need to confirm with the Board of Elections what I think to be ghost precincts. I also want to confirm I addressed ballot results listed as overvote, undervote or write-in correctly.
