---
title: "Draft story memo"
author: "Katelynn Winebrenner"
date: "June-1-2026"
format:
  html:
    theme: cosmo
---

katelynnwinebrenner\@gmail.com 240-362-4882 June 1, 2026

# 2024 Frederick County School Board Election

In the 2024 Frederick County School Board election, the Frederick County Democratic Party endorsed four candidates for three open seats. This decision caused splintering within the party that led to conservative **candidates backed by Moms for Liberty** to take two of the seats.

IDENTIFY CANDIDATES BACKED BY BLUE OR RED

<iframe title="Democrat-backed school board candidates splintered precinct votes in 2024" aria-label="Choropleth map" id="datawrapper-chart-S5pgh" src="https://datawrapper.dwcdn.net/S5pgh/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height data-external="1">

</iframe>

```{=html}
<script type="text/javascript">(function(){function e(){window.addEventListener(`message`,function(e){if(e.data[`datawrapper-height`]!==void 0){var t=document.querySelectorAll(`iframe`);for(var n in e.data[`datawrapper-height`])for(var r=0,i;i=t[r];r++)if(i.contentWindow===e.source){var a=e.data[`datawrapper-height`][n]+`px`;i.style.height=a}}})}e()})();</script>
```

<iframe title="Harris took Frederick County, but Trump wasn&#39;t far behind" aria-label="Choropleth map" id="datawrapper-chart-bakfo" src="https://datawrapper.dwcdn.net/bakfo/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height data-external="1">

</iframe>

## Relevance

This story is relevant now because no one has mapped out these votes yet to identify where the splintering happened, and there is another upcoming school board election for the other four seats.

TABLE

```{r setup, include=FALSE}
```

```{r}
#| include: false
library(tidyverse)
library(janitor)
library(dplyr)
library(sf)
library(tidyr)
```

```{r}
#| include: false

general1 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/GENERAL_11PG24_Prov_CVR_SBE_1.csv") |> 
  clean_names()

general2 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/master/refs/heads/main/data/GENERAL_11PG24_Prov_CVR_SBE.csv") |> 
  clean_names()

#the two general files from 2024 should be combined. they are the same sheet, one was just limited to 99,999 
general <- rbind(general1, general2)

```

```{r}
#| include: false
 general_cleaned <- general |>
  select(cast_vote_record, precinct, ballot_style, seat_one = board_of_education_500001671, seat_two = x13, seat_three = x14)

seat_one <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_one)
seat_two <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_two)
seat_three <- general_cleaned |> select(cast_vote_record, precinct, ballot_style, candidate = seat_three)

combined <- union(seat_one, seat_two,)
combined <-union(combined, seat_three)



```

```{r}
#| echo: true
precinct_vote_counts <- combined |>
   group_by(precinct, candidate)|>
   arrange(candidate)|>
  summarize(votes=n()) |>
  mutate(precinct= substring(precinct, 2))|>
  filter(
    !is.na(candidate),
    !candidate %in% c("undervote", "overvote", "or write-in:") #questionable
  ) |>
  group_by(precinct) |>
  arrange(desc(votes), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  ungroup()


precinct_vote_wide <- precinct_vote_counts |>
  pivot_wider(
    names_from = rank,
    values_from = c(candidate, votes),
  names_glue = "{.value}_{rank}")|>
 mutate(
    across(
      starts_with("candidate"),
      ~ trimws(gsub("\\s*\\([^)]*\\)", "",.x))
    ))

  #create a map to represent the percent of voters per precinct who voted for rep-backed candidates versus dem-backed candidates— easier to compare with kamala v trump

vote_percent_board <- precinct_vote_counts |>
  group_by(precinct)|>
  mutate(
    across(
            starts_with("candidate"),
               ~ trimws(gsub("\\s*\\([^)]*\\)", "",.x))),
    party =
          case_when(
              candidate == "Jaime Kiersten Brennan" ~ "*",
              candidate == "Colt Morningstar Black" ~ "*",
              candidate == "Janie Monier" ~ "^",
              candidate == "Josh Bokee" ~ "^",
              candidate == "Chad King Wilson, Sr." ~ "^",
              candidate == "Veronica D. Lowe"~ "^"))|>
        pivot_wider(
            names_from = rank,
            values_from = c(candidate, votes, party),
            names_glue = "{.value}_{rank}")|>
      mutate(
           total_votes = votes_1 + votes_2 + votes_3 + votes_4 + votes_5 + votes_6,
           percent_votes_1 = round((votes_1/total_votes)*100),
           percent_votes_2 = round((votes_2/total_votes)*100),
           percent_votes_3 = round((votes_3/total_votes)*100),
           percent_votes_4 = round((votes_4/total_votes)*100),
           percent_votes_5 = round((votes_5/total_votes)*100),
           percent_votes_6 = round((votes_6/total_votes)*100)
     
     )|>
    head()


```

Presidential election results by precinct

```{r}
#| echo: true
general_president_clean <- general|>
  select(precinct, president=president_and_vice_president_of_the_united_states_500001630)|>
  group_by(precinct, president)|>
     arrange(president)|>
  summarize(votes=n()) |>
  mutate(precinct= substring(precinct, 2))|>
  filter(
    !is.na(president),
    !president %in% c("undervote", "overvote", "or write-in:") #questionable
  ) |>
  group_by(precinct) |>
  arrange(desc(votes), .by_group = TRUE) |>
  mutate(rank = row_number()) |>
  ungroup()|>
  mutate(
    across(
            starts_with("president"),
               ~ trimws(gsub("\\s*\\([^)]*\\)", "",.x))))|>
  pivot_wider(
    names_from = rank,
    values_from = c(president, votes),
  names_glue = "{.value}_{rank}")|>
      mutate(
           total_votes = votes_1 + votes_2 + votes_3 + votes_4 + votes_5,
           percent_votes_1 = round((votes_1/total_votes)*100),
           percent_votes_2 = round((votes_2/total_votes)*100),
           percent_votes_3 = round((votes_3/total_votes)*100),
           percent_votes_4 = round((votes_4/total_votes)*100),
           percent_votes_5 = round((votes_5/total_votes)*100)
     )|>
  head()

```

## Audience

My target audience is Frederick county parents and candidates.

## Sources

Some people I could interview for this story would be: Frederick Democrats Chair \| Mari Lee \| (301) 898-2513 \| info\@frederickdemocrats.org Moms for Liberty Chapter Chair \| Tiffany Noble \| Tiffany.noble0820\@gmail.com Frederick County PTA President \| Najlaa Richard \| president\@frederickpta.org School board candidate Chad Wilson, Sr. (lost in 2024, running again this season)

## Potential impact

The potential impact of this story would be to reveal key areas where splintering occurred to inform candidates and voters where the discrepancies were to increase voter participation and civic engagement in these areas.

## Interest

This story was particularly surprising to me because it showcases the impact of partisanship within the county school board, which is technically considered nonpartisan. The extreme ideological differences between the candidates is also intriguing to me, particularly because the former Frederick County Moms of Liberty stepped down to run for a seat and won. I’m also interested in this story because it shows that, despite Frederick’s growth in recent years that have skewed its politics to the left, the conservative values in the county still remain strong.

## Previous coverage

From what I’ve seen, there has not been much coverage on this race in particular, other than generalized statewide coverage about the Democrats’ failures in school boards throughout the state from Maryland Matters.

Half of ‘extreme’ school board candidates targeted by Md. Democratic Party win their races - Maryland Matters Md. Democratic Party targets 'extreme candidates ' in local school board races - Maryland Matters

## Visuals

**RSW COMMENT; LET'S GET THAT MAP BUILT TODAY** I’d like to create a chloropleth map of precincts that display where the splintering occurred. Comparing and contrasting that map to a map of 2024 election results would help contextualize these areas as leaning left or right as well.

## To learn

As for other information still to gather, I need to do more research on the previous election members, why those seats were open and what the 2024 campaign season was like. I’d also like to research what kind of decisions or changes the new conservative school board members have made since being elected to office and what new candidates are saying about them.

## To do and expected completion

I would like to have this done around the time of the primary on June 23. So far, I’ve already organized the vote cast records by precinct and candidate. I still have to display that visually and do the 2024 presidential election analysis to identify key geographic areas. Interviews shouldn’t take very long to set up. I would also like to try to talk to some regular Frederick County parents, which could be a little more difficult as a newcomer.

## Current code

-30-
